%% ── Chapter 12: Conclusion ───────────────────────────────────
\chapter{Conclusion}
\label{ch:conclusion}

\section{Summary of Contributions}

This thesis presented \caesar{}, a co-evolutionary AI framework for
simultaneous attack simulation and adaptive defense in cybersecurity.
Six novel algorithmic contributions were introduced and evaluated:

\begin{enumerate}
  \item \tagan{} demonstrated that conditioning attack generation on the
    defense state produces more diverse, effective adversarial traffic than
    unconditional generation (RQ1 confirmed).
  \item \adpn{} achieved a neutralization rate of $91.9\% \pm 3.3\%$ and
    robustness score of $0.967 \pm 0.007$ through co-evolutionary training,
    statistically significant at $p = 0.002$ (Wilcoxon), outperforming all
    static baselines and SOTA methods under adversarial conditions (RQ2 confirmed).
  \item \tagraph{} enabled proactive defense in 18\% of episodes with 63.2\%
    prediction accuracy, demonstrating the value of temporal graph memory
    for anticipatory threat response (RQ3 confirmed).
  \item \mgae{} achieved complete evasion (98.5\% $\to$ 0\% detection rate)
    against RF and DT classifiers on real CICIDS2017 data through
    manifold-guided diffusion (RQ4 confirmed).
  \item \arl{} achieved 100\% autonomous healing success over 300 ticks
    with zero human escalations, validating the self-healing AI security
    concept (RQ5 confirmed).
\end{enumerate}

\section{Future Work}

\begin{description}
  \item[Real-world data integration] Extend to full CICIDS2017 and Bot-IoT
    datasets with real network captures.
  \item[Graph Neural Networks for TAG] Replace Markov-chain prediction with
    a GNN operating directly on the \tagraph{} structure.
  \item[LLM integration] Fine-tune domain-specific language models for
    realistic spear-phishing generation (Phase 2 extension).
  \item[Federated \caesar{}] Deploy the co-evolutionary loop across
    distributed network segments using federated learning.
  \item[Formal verification] Prove convergence properties of the
    co-evolutionary loop under specific game-theoretic assumptions.
  \item[Hardware deployment] Implement the \arl{} as a real-time network
    appliance using FPGA acceleration.
\end{description}

\section{Closing Remarks}

The arms race between attackers and defenders is fundamentally a
co-evolutionary process. Static defenses will always lag behind adaptive
adversaries. \caesar{} demonstrates that AI can be placed on both sides
of this race — not to make attacks easier, but to train defenses that are
robust enough to face the attacks of tomorrow, today.
